---
title: "Extended Figure 3"
author: "Eva Brigos"
date: "2026-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library("ggpubr")
library(forcats)
library(cowplot)
```

# Load the dataframes

```{r}
# Genetic Correlations
gcF <- read.csv("fertility-genetic-correlations.txt", sep = "\t")
gcL <- read.csv("longevity-genetic-correlations.txt", sep = "\t")

# Age at disease onset
ado <- read.csv("onset_age_62diseases_v17.01.2025_FinRegistryR11.tsv", sep = "\t")

# Metadata
metadata <- read.csv("domain_name_acronym_62diseases.csv", sep = ";")
```

Rename the data-frame columns:

```{r}
names(ado) <- c("Code", "ADO_5", "ADO_50", "ADO_95")
names(gcF) <- c("Code", "Intercept", "Intercept_SE", "GC", "GC_SE", "PValue")
names(gcL) <- c("Code", "Intercept", "Intercept_SE", "GC", "GC_SE", "PValue")
```

Create a new merged data-frame:

```{r}
# Merge gc with ado
gc_adoF <- merge(gcF, ado, by = "Code")
gc_adoL <- merge(gcL, ado, by = "Code")

# Merge to have GC information
df_F <- merge(gc_adoF, metadata, by = "Code")
df_L <- merge(gc_adoL, metadata, by = "Code")
```

# Select diseases based on significance threshold

```{r}
# Compute FDR
df_F$FDR <- p.adjust(df_F$PValue, method = "BH")
df_L$FDR <- p.adjust(df_L$PValue, method = "BH")

# Filter for significance
df2_F <- df_F %>% filter(PValue < 0.05)
df2_L <- df_L %>% filter(PValue < 0.05)
```

```{r}
df2_F$Correlation <- ifelse(df2_F$GC > 0, "Positive GC", "Negative GC")
table(df2_F$Correlation)

df2_L$Correlation <- ifelse(df2_L$GC > 0, "Positive GC", "Negative GC")
table(df2_L$Correlation)
```

# Correlation test

```{r}
# Change GCST010003 from 1.26 to 1.27 to avoid the exact same value with GCST90014023
# Change GCST003763 from 24.6 to 24.7 to avoid the exact same value with GCST90129439
df2_L[28, "ADO_5"]
df2_L[25, "ADO_5"]
df2_L[28, "ADO_5"] <- 1.27
df2_L[25, "ADO_5"] <- 24.7
```

Correlation test: 

```{r}
correlation_result_F <- cor.test(df2_F$GC, df2_F$ADO_5, method = "spearman")
print(correlation_result_F)
correlation_F <- correlation_result_F$estimate
p_value_F <- correlation_result_F$p.value

correlation_result_L <- cor.test(df2_L$GC, df2_L$ADO_5, method = "spearman")
print(correlation_result_L)
correlation_L <- correlation_result_L$estimate
p_value_L <- correlation_result_L$p.value
```

# Do the plot by disease domain

Create a dictionary to map the domain names with new shorter version:

```{r}
domain_dict <- c("Diseases of the circulatory system" = "Circulatory system",
                  "Diseases of the ear or mastoid process" = "Ear or mastoid process",
                  "Diseases of the genitourinary system" = "Genitourinary system",
                  "Diseases of the nervous system" = "Nervous system",
                  "Diseases of the skin" = "Diseases of the skin",
                  "Mental, behavioural or neurodevelopmental disorders" = "Mental, behavioural\nor neurodevelopmental",
                  "Diseases of the digestive system" = "Digestive system",
                  "Diseases of the eye and adnexa" = "Eye and adnexa",
                  "Diseases of the musculoskeletal system or connective tissue" = "Musculoskeletal system\nor connective tissue",
                  "Diseases of the respiratory system" = "Respiratory system",
                  "Endocrine, nutritional or metabolic diseases" = "Endocrine, nutritional\nor metabolic",
                  "Neoplasms" = "Neoplasms")

df_F$DomainShort <- domain_dict[df_F$Domain]
df2_F$DomainShort <- domain_dict[df2_F$Domain]

df_L$DomainShort <- domain_dict[df_L$Domain]
df2_L$DomainShort <- domain_dict[df2_L$Domain]
```

Dictionary for the colors:

```{r}
domcolors <- c(
  "Diseases of the skin" = "black",
  "Mental, behavioural\nor neurodevelopmental" = "#56B4E9",
  "Digestive system" ="#332288",
  "Eye and adnexa" = "#F0E442",
  "Respiratory system" = "#0072B2",
  "Nervous system" = "#D55E00",
  "Neoplasms" = "#CC79A7",
  "Endocrine, nutritional\nor metabolic" = "#E69F00",
  "Circulatory system" = "#882255",
  "Musculoskeletal system\nor connective tissue" = "#117733",
  "Genitourinary system" = "#009E73",
  "Ear or mastoid process" = "#DDCC77"
)
```

With the significant diseases FERTILITY:

```{r}
gc_domain_sig_F <- df2_F %>%
  mutate(class = fct_reorder(DomainShort, GC, .fun='median')) %>%
  ggplot( aes(x=reorder(DomainShort, GC), y=GC)) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "black", linewidth = 0.5) +
    geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = 0), fill = "#F8F8FF", alpha = 0.5) +
    geom_boxplot(aes(fill = DomainShort)) +
  geom_point(position = position_dodge(width = 0.75), size = 1.5, color = "black") +
  scale_fill_manual(values = domcolors) +
  theme_bw() +
  labs(y = expression(paste("Genetic correlation (", italic(r)[g], ")"))) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1,
                                            size = 7, family = "Helvetica", color = "black"),
        axis.text.y = element_text(size = 7, family = "Helvetica", color = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7, color = "black", margin = margin(r = 10), family = "Helvetica"),
        plot.margin = margin(5, 5, 5, 5),
        panel.grid = element_blank())

```

With the significant diseases LONGEVITY:

```{r}
gc_domain_sig_longevity <- df2_L %>%
  mutate(class = fct_reorder(DomainShort, GC, .fun='median')) %>%
  ggplot( aes(x=reorder(DomainShort, GC), y=GC)) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "black", linewidth = 0.5) +
    geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = 0), fill = "#F8F8FF", alpha = 0.5) +
    geom_boxplot(aes(fill = DomainShort)) +
  geom_point(position = position_dodge(width = 0.75), size = 1.5, color = "black") +
  scale_fill_manual(values = domcolors) +
  theme_bw() +
  labs(y = expression(paste("Genetic correlation (", italic(r)[g], ")"))) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1,
                                            size = 7, family = "Helvetica", color = "black"),
        axis.text.y = element_text(size = 7, family = "Helvetica", color = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7, color = "black", margin = margin(r = 10), family = "Helvetica"),
        plot.margin = margin(5, 5, 5, 5),
        panel.grid = element_blank())

```

Merge into one figure: 

```{r}
gc_domains_plot <- plot_grid(
  gc_domain_sig_longevity, gc_domain_sig_F,
  ncol = 1,
  labels = c("a", "b"),
  label_fontface = "bold",
  label_size = 7,
  align = "hv",
  rel_widths = c(1, 1)
)
```

Save the plot:

```{r}
ggsave("/extended_data_fig_2.png", gc_domains_plot, height = 180, width = 180, units = "mm", dpi = 400)
```

Vector file:

```{r}
ggsave(
  filename = "extended_data_fig_2.pdf",
  plot = gc_domains_plot,
  device = "pdf",
  width = 180,
  height = 180,
  units = "mm",
  useDingbats = FALSE
)
```

# Correlation between onset age and LDSC results

Add the correlation and significance columns FERTILITY:

```{r}
df_F$Correlation <- ifelse(df_F$GC > 0, "Positive GC", "Negative GC")
df_F$Significant <- ifelse(df_F$PValue < 0.05, "Significant", "Not Significant")
```

Do the plot:

```{r}
corr_gc_ado_fertility <- ggplot(df_F, aes(x = ADO_5, y = GC, color = Correlation)) +
  geom_point(alpha = 0.3, size = 2) +
  geom_point(data = df_F, aes(alpha = Significant), size = 2)+
  scale_color_manual(values = c("Positive GC" = "#023E8A", "Negative GC" = "#990000")) +
  geom_smooth(data = df2_F, method = "lm", se = TRUE, color = "black", size = 0.5) +
  theme_bw() +
  annotate("text", x = Inf, y = Inf,
           label = paste("R =", round(correlation_F, 2), "\nP =",
                         format(p_value_F, scientific = TRUE)), 
           hjust = 1.1, vjust = 2, size = 7/.pt, color = "black")
corr_gc_ado_fertility <- corr_gc_ado_fertility + theme(legend.position = "none",
                     axis.title.x = element_text(size = 7, margin = margin(t = 6), family = "Helvetica"),
                     axis.title.y = element_text(size = 7, margin = margin(r = 6), family = "Helvetica"),
                     axis.text.y = element_text(size = 7, family = "Helvetica"),
                     axis.text.x = element_text(size = 7, family = "Helvetica"),
                     plot.margin = margin(t = 5, r = 5, b = 5, l = 5),
                     axis.line = element_line(colour = "black"),
                     panel.grid = element_blank())
corr_gc_ado_fertility <- corr_gc_ado_fertility + ylab(expression(paste("Genetic correlation (", italic(r)[g], ")")))
corr_gc_ado_fertility <- corr_gc_ado_fertility + xlab("Age at disease onset")
```

Add the correlation and significance columns LONGEVITY:

```{r}
df_L$Correlation <- ifelse(df_L$GC > 0, "Positive GC", "Negative GC")
df_L$Significant <- ifelse(df_L$PValue < 0.05, "Significant", "Not Significant")
```

Do the plot:

```{r}
corr_gc_ado_longevity <- ggplot(df_L, aes(x = ADO_5, y = GC, color = Correlation)) +
  geom_point(alpha = 0.3, size = 2) +
  geom_point(data = df_L, aes(alpha = Significant), size = 2)+
  scale_color_manual(values = c("Positive GC" = "#023E8A", "Negative GC" = "#990000")) +
  geom_smooth(data = df2_L, method = "lm", se = TRUE, color = "black", size = 0.5) +
  theme_bw() +
  annotate("text", x = Inf, y = Inf,
           label = paste("R =", round(correlation_L, 2), "\nP =",
                         format(p_value_L, scientific = TRUE)), 
           hjust = 1.1, vjust = 2, size = 7/.pt, color = "black")
corr_gc_ado_longevity <- corr_gc_ado_longevity + theme(legend.position = "none",
                     axis.title.x = element_text(size = 7, margin = margin(t = 6), family = "Helvetica"),
                     axis.title.y = element_text(size = 7, margin = margin(r = 6), family = "Helvetica"),
                     axis.text.y = element_text(size = 7, family = "Helvetica"),
                     axis.text.x = element_text(size = 7, family = "Helvetica"),
                     plot.margin = margin(t = 5, r = 5, b = 5, l = 5),
                     axis.line = element_line(colour = "black"),
                     panel.grid = element_blank())
corr_gc_ado_longevity <- corr_gc_ado_longevity + ylab(expression(paste("Genetic correlation (", italic(r)[g], ")")))
corr_gc_ado_longevity <- corr_gc_ado_longevity + xlab("Age at disease onset")
```

Merge:

```{r}
corr_ado_plot <- plot_grid(
  corr_gc_ado_longevity, corr_gc_ado_fertility,
  ncol = 1,
  labels = c("a", "b"),
  label_fontface = "bold",
  label_size = 7,
  align = "hv",
  rel_widths = c(1, 1)
)
```

# Legend

```{r}
legend_plot <- ggplot(
  data.frame(
    group = c("Positive Pleiotropy", "Negative Pleiotropy"),
    x = 1, y = 1
  ),
  aes(x = x, y = y, color = group)
) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c(
      "Positive Pleiotropy" = "#023E8A",
      "Negative Pleiotropy"   = "#990000"
    )
  ) +
  guides(color = guide_legend(
    title = NULL,
    override.aes = list(shape = 16, size = 2.5)
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6, family = "Helvetica", color = "black"),
    legend.key.height = unit(0.60, "cm"),
    legend.key.width  = unit(0.20, "cm"),
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.box.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.margin = margin(2, 2, 2, 2)
  )

legend_grob <- get_legend(legend_plot)

corr_ado_plot_legend <- ggdraw(corr_ado_plot) +
  draw_grob(
    legend_grob,
    x = 0.8, y = 0.50,
    width = 0.20, height = 0.25
  )

corr_ado_plot_legend
```

# Save

```{r}
ggsave2("extended_data_fig_3.png", corr_ado_plot_legend, height = 170, width = 180, units = "mm", dpi = 400)
```

Vector file:

```{r}
ggsave(
  filename = "extended_data_fig_3.pdf",
  plot = corr_ado_plot_legend,
  device = "pdf",
  width = 180,
  height = 170,
  units = "mm",
  useDingbats = FALSE
)
```
