---
title: "Demography Extended Fig"
author: "Eva Brigos"
date: "2026-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}
library(dplyr)
library(purrr)
library(ggplot2)
library(ggbreak)
library(tidyr)
library("ggpubr")
library(forcats)
library(cowplot)
```

# Common Inputs

```{r}
# Age at disease onset
ado <- read.csv("onset_age_62diseases_v17.01.2025_FinRegistryR11.tsv",sep = "\t")
colnames(ado)[1] <- "Code"

# Load the metadata file
metadata <- read.csv("metadata_62complexdiseases.csv", sep = ";")
meta <- metadata %>% select(Code, Percentage_UKBiobank_Participants, Disease_Name)

# Load the N cases file
ncases <- read.csv("samplesize_decile10_62diseases.csv", sep = ",", header = TRUE)
```

# Number of children

Define the directory containing the files:

```{r}
file_directory <- "nchildren_rmarkdown_analysis_v16.03.2025"
```

List all 62 files:

```{r}
file_list <- list.files(path = file_directory, pattern = "\\_mean_sd_all.tsv$", recursive = TRUE, full.names = TRUE)
```

Initialize an empty data frame to store the results:

```{r}
result_df <- data.frame(Disease = character(), Mean_Case = numeric(), Mean_Control = numeric(), stringsAsFactors = FALSE)
```

Iterate through each file to extract required information:

```{r}
disease_array <- c()
cases_mean_array <- c()
controls_mean_array <- c()
cases_sd_array <- c()
controls_sd_array <- c()

for (file_path in file_list) {
  data <- read.csv(file_path, sep = "\t")
  disease_code <- data$Disease[1]
  mean_case <- data$mean[data$Status == "Decil 10, cases"]
  mean_control <- data$mean[data$Status == "Decil 10, controls"]
  sd_case <- data$sd[data$Status == "Decil 10, cases"]
  sd_control <- data$sd[data$Status == "Decil 10, controls"]
  disease_array <- c(disease_array, disease_code)
  cases_mean_array <- c(cases_mean_array, mean_case)
  controls_mean_array <- c(controls_mean_array, mean_control)
  cases_sd_array <- c(cases_sd_array, sd_case)
  controls_sd_array <- c(controls_sd_array, sd_control)
}
```

Create the new data-frame:

```{r}
result_fertility <- data.frame(Code = disease_array, Mean_Case = cases_mean_array,
                        Mean_Control = controls_mean_array,
                        SD_Case = cases_sd_array,
                        SD_Control = controls_sd_array
                        )
```

# Merge with N cases

```{r}
result_fertility <- result_fertility %>%
  left_join(ncases, by = "Code")
```

# Significance test

```{r}
result_fertility <- result_fertility %>%
  mutate(
    SE_cases    = ifelse(!is.na(SD_Case) & !is.na(NCases) & NCases > 0, SD_Case / sqrt(NCases), NA_real_),
    SE_controls = ifelse(!is.na(SD_Control) & !is.na(NControls) & NControls > 0, SD_Control / sqrt(NControls), NA_real_),
    Diff = Mean_Control - Mean_Case,
    SE_Diff = sqrt(SE_controls^2 + SE_cases^2),
    t_stat = Diff / SE_Diff,
    df_welch = (SE_controls^2 + SE_cases^2)^2 /
      ((SE_controls^4) / (NControls - 1) + (SE_cases^4) / (NCases - 1)),
    p_value = 2 * pt(-abs(t_stat), df = df_welch),
    Mean_non_significant = case_when(
      !is.na(p_value) & p_value < 0.05 ~ "significant",
      !is.na(p_value) ~ "non-significant",
      TRUE ~ NA_character_
    )
  )
```

# ADO

```{r}
df_plot_fertility <- ado %>%
  left_join(result_fertility, by = "Code") %>%
  left_join(meta, by = "Code")
```

```{r}
df_plot_fertility <- df_plot_fertility %>%
  mutate(Disease_Name = ifelse(is.na(Disease_Name) | Disease_Name == "", Code, Disease_Name))

df_plot_fertility$Disease_Name <- factor(df_plot_fertility$Disease_Name, levels = unique(df_plot_fertility$Disease_Name[order(df_plot_fertility$ADO_5, decreasing = TRUE)]))
```

# Color

```{r}
df_plot_fertility <- df_plot_fertility %>%
  mutate(
    point_color = case_when(
      Mean_non_significant == "non-significant" ~ "Non-significant",
      Mean_non_significant == "significant" & Mean_Case < Mean_Control ~ "Case < Control",
      Mean_non_significant == "significant" & Mean_Case >= Mean_Control ~ "Case > Control",
      TRUE ~ NA_character_
    )
  )
```

# Limits

```{r}
fertility_dropNA <- na.omit(df_plot_fertility)
xvals <- c(fertility_dropNA$Mean_Case, fertility_dropNA$Mean_Control)
xvals <- xvals[is.finite(xvals)]
x_min <- min(xvals)
x_max <- max(xvals)
pad <- 0.02 * (x_max - x_min)
if (!is.finite(pad) || pad == 0) pad <- 0.01
```

```{r}
df_plot_fertility$Difference <- df_plot_fertility$Mean_Control - df_plot_fertility$Mean_Case
```

Plot:

```{r}
# 1)
df_plot_fertility2 <- df_plot_fertility %>%
  mutate(
    diff = Mean_Control - Mean_Case,
    diff_abs = abs(diff)
  )

df_plot_fertility2_dropNA <- na.omit(df_plot_fertility2)

# 2)
pad <- if (!exists("pad")) 0 else pad
x_max <- max(df_plot_fertility2_dropNA$diff_abs, na.rm = TRUE)
x_lim <- x_max + pad

# 3)
subplot1_option3 <- ggplot(df_plot_fertility2, aes(y = Disease_Name)) +
  geom_point(
    aes(x = diff_abs, color = point_color),
    shape = 16, size = 2.5, alpha = 1, na.rm = TRUE
  ) +
  scale_color_manual(values = c(
    "Non-significant" = "#E3E3E3D9",
    "Case < Control"  = "#EE3B3BD9",
    "Case > Control"  = "#000000D9"
  ), na.value = NA) +
  scale_y_discrete(drop = FALSE) +
  xlab("Number of children") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.y = element_text(vjust = 0.3, size = 6, family = "Helvetica", color = "black"),
    axis.title.x = element_text(family = "Helvetica", size = 7, margin = margin(t = 3), color = "black"),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 6, family = "Helvetica", color = "black", margin = margin(t = 2)),
    plot.margin = margin(t = 10, r = 1, b = 1, l = 1),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    plot.title.position = "plot",
    plot.title = element_text(family = "Helvetica", hjust = 0.82, vjust = 0, size = 7, color = "black", face = "bold")
  )

subplot1_option3
```


# Childlessness

```{r}
result_childless <- read.csv("childlessness_summary_by_disease.tsv", sep = "\t")
```

# Merge with N cases

```{r}
result_childless <- result_childless %>%
  left_join(ncases, by = "Code")
```

# Significance test

```{r}
result_childless <- result_childless %>%
  mutate(
    p_case_raw    = Pct_NoChildren_Case,
    p_control_raw = Pct_NoChildren_Control,
    p_case = if_else(!is.na(p_case_raw) & p_case_raw > 1, p_case_raw / 100, p_case_raw),
    p_control = if_else(!is.na(p_control_raw) & p_control_raw > 1, p_control_raw / 100, p_control_raw),
    p_case    = if_else(!is.na(p_case),    pmin(pmax(p_case, 0), 1), NA_real_),
    p_control = if_else(!is.na(p_control), pmin(pmax(p_control, 0), 1), NA_real_),
    x_case    = if_else(!is.na(p_case)    & !is.na(NCases)    & NCases    > 0, round(p_case    * NCases),    NA_real_),
    x_control = if_else(!is.na(p_control) & !is.na(NControls) & NControls > 0, round(p_control * NControls), NA_real_),
    Diff = p_control - p_case,
    p_pool = if_else(
      !is.na(x_case) & !is.na(x_control) &
        !is.na(NCases) & !is.na(NControls) &
        NCases > 0 & NControls > 0,
      (x_case + x_control) / (NCases + NControls),
      NA_real_
    ),
    SE_Diff = if_else(
      !is.na(p_pool) & p_pool >= 0 & p_pool <= 1 &
        !is.na(NCases) & !is.na(NControls) &
        NCases > 0 & NControls > 0,
      sqrt(p_pool * (1 - p_pool) * (1 / NControls + 1 / NCases)),
      NA_real_
    ),
    z_stat  = if_else(!is.na(SE_Diff) & SE_Diff > 0, Diff / SE_Diff, NA_real_),
    p_value = if_else(!is.na(z_stat), 2 * pnorm(-abs(z_stat)), NA_real_),
    Mean_non_significant = case_when(
      !is.na(p_value) & p_value < 0.05 ~ "significant",
      !is.na(p_value) ~ "non-significant",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-p_case_raw, -p_control_raw)
```

# ADO

```{r}
df_plot_childless <- ado %>%
  left_join(result_childless, by = "Code") %>%
  left_join(meta, by = "Code")
```

```{r}
df_plot_childless <- df_plot_childless %>%
  mutate(Disease_Name = ifelse(is.na(Disease_Name) | Disease_Name == "", Code, Disease_Name))

df_plot_childless$Disease_Name <- factor(df_plot_childless$Disease_Name, levels = unique(df_plot_childless$Disease_Name[order(df_plot_childless$ADO_5, decreasing = TRUE)]))
```

# Color

```{r}
df_plot_childless <- df_plot_childless %>%
  mutate(
    point_color = case_when(
      Mean_non_significant == "non-significant" ~ "Non-significant",
      Mean_non_significant == "significant" & p_case < p_control ~ "Case < Control",
      Mean_non_significant == "significant" & p_case >= p_control ~ "Case > Control",
      TRUE ~ NA_character_
    )
  )
```

# Limits

```{r}
childless_dropNA <- na.omit(df_plot_childless)
xvals <- c(childless_dropNA$p_case, childless_dropNA$p_control)
xvals <- xvals[is.finite(xvals)]
x_min <- min(xvals)
x_max <- max(xvals)
pad <- 0.02 * (x_max - x_min)
if (!is.finite(pad) || pad == 0) pad <- 0.01
```

Plot:

```{r}
# 1)
df_plot_childless2 <- df_plot_childless %>%
  mutate(
    diff = Pct_NoChildren_Control - Pct_NoChildren_Case,
    diff_abs = abs(diff)
  )

df_plot_childless2_dropNA <- na.omit(df_plot_childless2)

# 2)
pad <- if (!exists("pad")) 0 else pad
x_max <- max(df_plot_childless2_dropNA$diff_abs, na.rm = TRUE)
x_lim <- x_max + pad

# 3)
subplot2_option3 <- ggplot(df_plot_childless2, aes(y = Disease_Name)) +
  geom_point(
    aes(x = diff_abs, color = point_color),
    shape = 16, size = 2.5, alpha = 1, na.rm = TRUE
  ) +
  scale_color_manual(values = c(
    "Non-significant" = "#E3E3E3D9",
    "Case < Control"  = "#000000D9",
    "Case > Control"  = "#EE3B3BD9"
  ), na.value = NA) +
  scale_y_discrete(drop = FALSE) +
  coord_cartesian(xlim = c(0, x_lim)) +
  xlab("Childlessness %") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.x = element_text(family = "Helvetica", size = 7, margin = margin(t = 3), color = "black"),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 6, family = "Helvetica", color = "black", margin = margin(t = 2)),
    plot.margin = margin(t = 10, r = 1, b = 1, l = 1),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    plot.title.position = "plot",
    plot.title = element_text(family = "Helvetica", hjust = 0.82, vjust = 0, size = 7, color = "black", face = "bold")
  )

subplot2_option3
```

# Age at first birth (AFB)

```{r}
result_afb <- read.csv("afb_summary_by_disease.tsv", sep = "\t")
```

# Merge with N cases

```{r}
result_afb <- result_afb %>%
  left_join(ncases, by = "Code")
```

# Significance test

```{r}
result_afb <- result_afb %>%
  mutate(
    SE_cases    = ifelse(!is.na(SD_Case) & !is.na(NCases) & NCases > 0, SD_Case / sqrt(NCases), NA_real_),
    SE_controls = ifelse(!is.na(SD_Control) & !is.na(NControls) & NControls > 0, SD_Control / sqrt(NControls), NA_real_),
    Diff = Mean_Control - Mean_Case,
    SE_Diff = sqrt(SE_controls^2 + SE_cases^2),
    t_stat = Diff / SE_Diff,
    df_welch = (SE_controls^2 + SE_cases^2)^2 /
      ((SE_controls^4) / (NControls - 1) + (SE_cases^4) / (NCases - 1)),
    p_value = 2 * pt(-abs(t_stat), df = df_welch),
    Mean_non_significant = case_when(
      !is.na(p_value) & p_value < 0.05 ~ "significant",
      !is.na(p_value) ~ "non-significant",
      TRUE ~ NA_character_
    )
  )
```

# ADO

```{r}
df_plot_afb <- ado %>%
  left_join(result_afb, by = "Code") %>%
  left_join(meta, by = "Code")
```

```{r}
df_plot_afb <- df_plot_afb %>%
  mutate(Disease_Name = ifelse(is.na(Disease_Name) | Disease_Name == "", Code, Disease_Name))

df_plot_afb$Disease_Name <- factor(df_plot_afb$Disease_Name, levels = unique(df_plot_afb$Disease_Name[order(df_plot_afb$ADO_5, decreasing = TRUE)]))
```

# Color

```{r}
df_plot_afb <- df_plot_afb %>%
  mutate(
    point_color = case_when(
      Mean_non_significant == "non-significant" ~ "Non-significant",
      Mean_non_significant == "significant" & Mean_Case < Mean_Control ~ "Case < Control",
      Mean_non_significant == "significant" & Mean_Case >= Mean_Control ~ "Case > Control",
      TRUE ~ NA_character_
    )
  )
```

# Limits

```{r}
afb_dropNA <- na.omit(df_plot_afb)
xvals <- c(afb_dropNA$Mean_Case, afb_dropNA$Mean_Control)
xvals <- xvals[is.finite(xvals)]
x_min <- min(xvals)
x_max <- max(xvals)
pad <- 0.02 * (x_max - x_min)
if (!is.finite(pad) || pad == 0) pad <- 0.01
```

```{r}
df_plot_afb$Difference <- df_plot_afb$Mean_Control - df_plot_afb$Mean_Case
```

Mean difference:

```{r}
mean(afb_dropNA$Diff[afb_dropNA$point_color == "Case < Control"])
sd(afb_dropNA$Diff[afb_dropNA$point_color == "Case < Control"])
```

Plot:

```{r}
# 1)
df_plot_afb2 <- df_plot_afb %>%
  mutate(
    diff = Mean_Control - Mean_Case,
    diff_abs = abs(diff)
  )

df_plot_afb2_dropNA <- na.omit(df_plot_afb2)

# 2)
pad <- if (!exists("pad")) 0 else pad
x_max <- max(df_plot_afb2_dropNA$diff_abs, na.rm = TRUE)
x_lim <- x_max + pad

# 3)
subplot3_option3 <- ggplot(df_plot_afb2, aes(y = Disease_Name)) +
  geom_point(
    aes(x = diff_abs, color = point_color),
    shape = 16, size = 2.5, alpha = 1, na.rm = TRUE
  ) +
  scale_color_manual(values = c(
    "Non-significant" = "#E3E3E3D9",
    "Case > Control"  = "#EE3B3BD9",
    "Case < Control"  = "#000000D9"
  ), na.value = NA) +
  scale_y_discrete(drop = FALSE) +
  coord_cartesian(xlim = c(0, x_lim)) +
  xlab("Age at first birth") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.x = element_text(family = "Helvetica", size = 7, margin = margin(t = 3), color = "black"),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 6, family = "Helvetica", color = "black", margin = margin(t = 2)),
    plot.margin = margin(t = 10, r = 1, b = 1, l = 1),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    plot.title.position = "plot",
    plot.title = element_text(family = "Helvetica", hjust = 0.82, vjust = 0, size = 7, color = "black", face = "bold")
  )

subplot3_option3
```

# Onset age

Create a dummy variable for the y-axis:

```{r}
onset <- df_plot_fertility2 %>%
  mutate(dummy = 1)
```

Order by onset age:

```{r}
onset$Disease_Name <- factor(onset$Disease_Name, levels = unique(onset$Disease_Name[order(onset$ADO_5, decreasing = TRUE)]))
onset <- onset[order(onset$ADO_5),]
```

Get the lables to plot:

```{r}
values_onset <- as.array(round(onset$ADO_5, 0))
names_disease <- as.array(onset$Disease_Name)
```

Plot:

```{r}
subplot_ado <- ggplot(onset, aes(y = Disease_Name, x = dummy, fill = ADO_5)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_gradient(low = "#FEE08B", high = "#B7E7A7") +
  scale_x_discrete(expand = expansion(add = c(0, 0))) +
  scale_y_discrete(breaks = onset$Disease_Name, labels = values_onset) +
  theme_bw() +
  xlab("Onset\nAge") +
  theme(
    legend.position = "none",
    axis.title = element_blank(),
    axis.title.x = element_text(size = 6, margin = margin(b = 1.5), color = "black", family = "Helvetica"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(t = 10, r = 5, b = 1, l = 1)
    )

subplot_ado
```

Add ticks:

```{r}
custom_ticks <- data.frame(
  y = c(61, 48, 38, 28, 20, 11, 5, 1.5),
  label = c(1, 10, 20, 30, 40, 45, 50, 65)
)

subplot_ado <- subplot_ado +
  geom_text(data = custom_ticks,
          aes(x = 1, y = y, label = label),
          inherit.aes = FALSE,
          size = 2, family = "Helvetica", color = "black", fontface = "bold")

subplot_ado
```

# UK Biobank Percentage

```{r}
subplot_ukb <- ggplot(df_plot_afb2, aes(y = Disease_Name, x = Percentage_UKBiobank_Participants)) +
  geom_col(fill = "#000000D9", width = 0.9, na.rm = TRUE) +
  scale_y_discrete(drop = FALSE) +
  xlab("% UK\nBiobank") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(family = "Helvetica", size = 6, margin = margin(t = 3), color = "black"),
    axis.text.x = element_text(size = 5, family = "Helvetica", color = "black", margin = margin(t = 2)),
    axis.title = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid   = element_blank(),
    axis.line    = element_line(color = "black"),
    plot.margin  = margin(t = 10, r = 5, b = 1, l = 1)
  )

subplot_ukb
```

# Merge figure

```{r}
figure1 <- plot_grid(subplot1_option3, subplot2_option3, subplot3_option3, subplot_ado,
  ncol = 4,
  labels = c("a", "b", "c", "d"),
  label_fontface = "bold",
  label_fontfamily = "Helvetica",
  label_size = 7,
  align = "h",   # Align vertically
  rel_heights = c(1, 1),
  rel_widths = c(1.95, 1.1, 1.1, 0.15)
)

figure1
```

# Legend

```{r}
legend_plot <- ggplot(
  data.frame(
    group = c("Undiagnosed have\nhigher fertility", "Diagnosed have\nhigher fertility"),
    x = 1, y = 1
  ),
  aes(x = x, y = y, color = group)
) +
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(
      "Undiagnosed have\nhigher fertility" = "#EE3B3BD9",  # red
      "Diagnosed have\nhigher fertility"   = "#000000D9"   # black
    )
  ) +
  guides(color = guide_legend(
    title = NULL,
    override.aes = list(shape = 16, size = 1.5)
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6, family = "Helvetica", color = "black"),
    legend.key.height = unit(0.60, "cm"),
    legend.key.width  = unit(0.35, "cm"),
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.box.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.margin = margin(3, 3, 3, 3)
  )

legend_grob <- get_legend(legend_plot)

figure1_with_legend <- ggdraw(figure1) +
  draw_grob(
    legend_grob,
    x = 0.27, y = 0.01,
    width = 0.20, height = 0.25
  )

figure1_with_legend
```

Save:

```{r}
ggsave("extended_data_fig_7.png",
       figure1_with_legend, 
       height = 180, 
       width = 180, 
       units = "mm",
       dpi = 400)
```

Vector file:

```{r}
ggsave(
  filename = "extended_data_fig_7.pdf",
  plot = figure1_with_legend,
  device = "pdf",
  width = 180,
  height = 180,
  units = "mm",
  useDingbats = FALSE
)
```
