---
title: "Functional Consequences: extended figures 4 & 5"
author: "Eva Brigos"
date: "2025-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library("ggpubr")
library(forcats)
library(cowplot)
library(ggrepel)
```

# Load the dataframes

```{r}
# Background set
vep_output <- read.csv("common_SNPs_62diseases_background_set.vep.txt", sep = "\t")
names(vep_output)[1] <- c("rsid")

# Longevity pleiotropies
longevity_pleios <- read.csv("plifespan_clumped_pleiotropies_62Diseases_commonSNPs_conjfdr005.csv", sep = ",")

# Fertility pleiotropies
fertility_pleios <- read.csv("fertility_clumped_pleiotropies_62Diseases_commonSNPs_conjfdr005.csv", sep = ",")
```

# Filter the background set

```{r}
# Column CANONICAL == YES
vep_output <- vep_output %>%
  filter(CANONICAL == "YES")

# New colum with rsid + Gene
vep_output <- vep_output %>%
  mutate(rsid_gene = paste(rsid, Gene, sep = "_"))

# Remove duplicates by rsid_gene column
vep_output <- vep_output %>%
  distinct(rsid_gene, .keep_all = TRUE)

# In the normal set we have 36 types of consequences, many of them with almost
# 0% frequency. I am going to group some of them into the bigger categories.
vep_output <- vep_output %>%
  mutate(Consequence_mod = if_else(
    Consequence == "intron_variant,non_coding_transcript_variant",
    Consequence,
    sub(",.*", "", Consequence)
  ))
```

Count the number of occurrences by functional consequence category:

```{r}
consequence_counts <- vep_output %>%
  group_by(Consequence_mod) %>%
  summarise(count = n()) %>%
  mutate(percentage = round((count / sum(count)) * 100, 1))
```

Change the name of Consequence_mod categories if percentage < 0.1:

```{r}
consequence_counts <- consequence_counts %>%
  mutate(Consequence_mod = if_else(
    percentage < 0.1,
    "Other",
    Consequence_mod
  ))
```

Aggregate the "Other" category:

```{r}
consequence_counts2 <- consequence_counts %>%
  group_by(Consequence_mod) %>%
  summarise(count = sum(count)) %>%
  mutate(percentage = round((count / sum(count)) * 100, 1))
```

Create the text to plot:

```{r}
consequence_counts2 <- consequence_counts2 %>%
  mutate(label_text = paste(Consequence_mod, " (", percentage, "%)", sep = "")) %>%
  mutate(text = paste(percentage, "%", sep = ""))

# Order descending
consequence_counts2 <- consequence_counts2 %>%
  arrange(desc(count))

# Put NA values in text if percentage < 12
consequence_counts2$text[consequence_counts2$percentage < 12] <- NA

```

Make shorter label names to facilitate plotting:

```{r}
consequence_counts2$label_text <- ifelse(consequence_counts2$label_text == "intron_variant,non_coding_transcript_variant (14.7%)",
                                              "intron_variant,non_coding (14.7%)", consequence_counts2$label_text)
consequence_counts2$label_text <- ifelse(consequence_counts2$label_text == "non_coding_transcript_exon_variant (1.2%)",
                                              "non_coding_transcript_exon (1.2%)", consequence_counts2$label_text)
consequence_counts2$label_text <- ifelse(consequence_counts2$label_text == "splice_polypyrimidine_tract_variant (0.1%)",
                                              "splice_polypyrimidine_tract (0.1%)", consequence_counts2$label_text)

consequence_counts2$category <- factor(consequence_counts2$label_text, levels = consequence_counts2$label_text)
```

Plot:

```{r}
par(family = "Helvetica")
```

Count the number of occurrences by biotype category:

```{r}
biotype_counts <- vep_output %>%
  group_by(BIOTYPE) %>%
  summarise(count = n()) %>%
  mutate(percentage = round((count / sum(count)) * 100, 1))
```

Change the name of biotype categories if percentage < 0.5:

```{r}
biotype_counts <- biotype_counts %>%
  mutate(BIOTYPE = if_else(
    percentage < 0.5,
    "Other",
    BIOTYPE
  ))
```

Aggregate the "Other" category:

```{r}
biotype_counts2 <- biotype_counts %>%
  group_by(BIOTYPE) %>%
  summarise(count = sum(count)) %>%
  mutate(percentage = round((count / sum(count)) * 100, 1))
```

Create the text to plot:

```{r}
biotype_counts2 <- biotype_counts2 %>%
  mutate(label_text = paste(BIOTYPE, " (", percentage, "%)", sep = "")) %>%
  mutate(text = paste(percentage, "%", sep = ""))

# Order descending
biotype_counts2 <- biotype_counts2 %>%
  arrange(desc(count))

# Put NA values in text if percentage < 5
biotype_counts2$text[biotype_counts2$percentage < 5] <- NA
biotype_counts2$category <- factor(biotype_counts2$label_text, levels = biotype_counts2$label_text)
```

# Do the analysis for the two life-history traits

Get the metadata:

```{r}
domains <- read.csv("domain_name_acronym_62diseases.csv", sep = ";")
names(domains)[1] <- c("disease_code")
```

Slice the background data-frame with relevant columns:

```{r}
info_background <- vep_output %>%
  select(rsid, Consequence, BIOTYPE, SYMBOL, Gene)
```

## Longevity

Merge the pleiotropy results with the background set. If an RSID has two or more
consequences (due to different gene isoforms for example), then all of them will
appear in the data-frame.

```{r}
longevity <- merge(longevity_pleios, info_background, by = "rsid")
```

### Longevity by domain

Merge with the domains:

```{r}
longevity <- merge(longevity, domains, by = "disease_code")
```

Create a dictionary to map the domain names with new shorter version:

```{r}
domain_dict <- c("Diseases of the circulatory system" = "Circulatory system",
                  "Diseases of the ear or mastoid process" = "Ear or mastoid process",
                  "Diseases of the genitourinary system" = "Genitourinary system",
                  "Diseases of the nervous system" = "Nervous system",
                  "Diseases of the skin" = "Diseases of the skin",
                  "Mental, behavioural or neurodevelopmental disorders" = "Mental, behavioural\nor neurodevelopmental",
                  "Diseases of the digestive system" = "Digestive system",
                  "Diseases of the eye and adnexa" = "Eye and adnexa",
                  "Diseases of the musculoskeletal system or connective tissue" = "Musculoskeletal system\nor connective tissue",
                  "Diseases of the respiratory system" = "Respiratory system",
                  "Endocrine, nutritional or metabolic diseases" = "Endocrine, nutritional\nor metabolic",
                  "Neoplasms" = "Neoplasms")

longevity$DomainShort <- domain_dict[longevity$Domain]
```

Create new data-frame centered in the disease domains:

```{r}
longevity_domain <- longevity %>%
  group_by(DomainShort) %>%
  summarize(total_positive = sum(pleiotropy_sign == "positive"),
            total_negative = sum(pleiotropy_sign == "negative")) %>%
  mutate(total = total_positive + total_negative,
         proportion_positive = total_positive / total,
         proportion_negative = total_negative / total)
```

Format the data for plotting:

```{r}
longevity_domain <- longevity_domain %>%
  arrange(desc(total))
longevity_domain$order <- c(1:11)

longevity_domain_long <- longevity_domain %>%
  pivot_longer(cols = starts_with("total"), names_to = "type", values_to = "count") %>%
  mutate(type = recode(type, total_positive = "Positive", total_negative = "Negative"))

longevity_domain_long <- longevity_domain_long[longevity_domain_long$type != "total", ]
```

Plot **number of pleiotropies by domain**:

```{r}
L1 <- ggplot(longevity_domain_long, aes(x = reorder(DomainShort, order),
                                            y = count,
                                            fill = type)) +
  geom_bar(stat = "identity") +
  labs(x = "Disease domain", y = "Number of pleiotropies") +
  scale_fill_manual(values = c("#990000CC", "#023E8ACC")) +
  theme_bw()+
  theme(text = element_text(size = 7, family = "Helvetica", color = "black"),
        axis.text.x = element_text(angle = 60, hjust = 1, size = 5),
        axis.text.y = element_text(size = 7),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7, margin = margin(0,10,0,0)),
        plot.margin = margin(10,10,2,10),
        panel.grid = element_blank())

pos_dom1 <- as.numeric(longevity_domain_long[1,"count"])
neg_dom1 <- as.numeric(longevity_domain_long[2,"count"])
pos_dom2 <- as.numeric(longevity_domain_long[3,"count"])
neg_dom2 <- as.numeric(longevity_domain_long[4,"count"])
pos_dom3 <- as.numeric(longevity_domain_long[5,"count"])
neg_dom3 <- as.numeric(longevity_domain_long[6,"count"])
pos_dom4 <- as.numeric(longevity_domain_long[7,"count"])
neg_dom4 <- as.numeric(longevity_domain_long[8,"count"])
pos_dom5 <- as.numeric(longevity_domain_long[9,"count"])
neg_dom5 <- as.numeric(longevity_domain_long[10,"count"])
pos_dom6 <- as.numeric(longevity_domain_long[11,"count"])
neg_dom6 <- as.numeric(longevity_domain_long[12,"count"])
pos_dom7 <- as.numeric(longevity_domain_long[13,"count"])
neg_dom7 <- as.numeric(longevity_domain_long[14,"count"])
pos_dom8 <- as.numeric(longevity_domain_long[15,"count"])
neg_dom8 <- as.numeric(longevity_domain_long[16,"count"])
pos_dom9 <- as.numeric(longevity_domain_long[17,"count"])
neg_dom9 <- as.numeric(longevity_domain_long[18,"count"])
pos_dom10 <- as.numeric(longevity_domain_long[19,"count"])
neg_dom10 <- as.numeric(longevity_domain_long[20,"count"])
pos_dom11 <- as.numeric(longevity_domain_long[21,"count"])
neg_dom11 <- as.numeric(longevity_domain_long[22,"count"])

for (i in 1:11) {
  pos_name <- paste("pos_dom", i, sep = "")
  neg_name <- paste("neg_dom", i, sep = "")
  pos <- get(pos_name)
  neg <- get(neg_name)
  if (!(pos == 0)) {
    L1 <- L1 + annotate("text", x = i, y = 15, label = as.character(pos), size = 5/.pt, color = "black")
  }
  if (!(neg == 0)) {
    L1 <- L1 + annotate("text", x = i, y = pos+neg+15, label = as.character(neg), size = 5/.pt, color = "black")
  }
}
```

### Longevity by functional consequence

Create new data-frame centered in the functional consequence:

```{r}
longevity_consequence <- longevity %>%
  group_by(Consequence) %>%
  summarize(total_positive = sum(pleiotropy_sign == "positive"),
            total_negative = sum(pleiotropy_sign == "negative")) %>%
  mutate(total = total_positive + total_negative,
         proportion_positive = total_positive / total,
         proportion_negative = total_negative / total)
```

Change the names for a shorter version:

```{r}
conseq_dict <- c("non_coding_transcript_exon_variant" = "non_coding_transcript\nexon_variant",
                  "splice_donor_region_variant,intron_variant" = "splice_donor_region_variant\nintron_variant",
                  "intron_variant,non_coding_transcript_variant" = "intron_variant\nnon_coding_transcript",
                  "splice_region_variant,splice_polypyrimidine_tract_variant,intron_variant" = "splice_region_variant\nsplice_polypyrimidine_tract\nintron_variant")

longevity_consequence <- longevity_consequence %>%
  mutate(
    Consequence = if_else(
      Consequence %in% names(conseq_dict),
      conseq_dict[Consequence],
      Consequence
    )
  )
```

Format the data for plotting:

```{r}
longevity_consequence <- longevity_consequence %>%
  arrange(desc(total))
longevity_consequence$order <- c(1:12)

longevity_consequence_long <- longevity_consequence %>%
  pivot_longer(cols = starts_with("total"), names_to = "type", values_to = "count") %>%
  mutate(type = recode(type, total_positive = "Positive", total_negative = "Negative"))

longevity_consequence_long <- longevity_consequence_long[longevity_consequence_long$type != "total", ]
```

Plot **number of pleiotropies by consequence**:

```{r}
L3 <- ggplot(longevity_consequence_long, aes(x = reorder(Consequence, order),
                                            y = count,
                                            fill = type)) +
  geom_bar(stat = "identity") +
  labs(x = "Functional consequence", y = "Number of pleiotropies") +
  scale_fill_manual(values = c("#990000CC", "#023E8ACC")) +
  theme_bw()+
  theme(text = element_text(size = 7, family = "Helvetica", color = "black"),
        axis.text.x = element_text(angle = 60, hjust = 1, size = 5),
        axis.text.y = element_text(size = 7),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7, margin = margin(0,10,0,0)),
        plot.margin = margin(10,10,2,10),
        panel.grid = element_blank())

pos_dom1 <- as.numeric(longevity_consequence_long[1,"count"])
neg_dom1 <- as.numeric(longevity_consequence_long[2,"count"])
pos_dom2 <- as.numeric(longevity_consequence_long[3,"count"])
neg_dom2 <- as.numeric(longevity_consequence_long[4,"count"])
pos_dom3 <- as.numeric(longevity_consequence_long[5,"count"])
neg_dom3 <- as.numeric(longevity_consequence_long[6,"count"])
pos_dom4 <- as.numeric(longevity_consequence_long[7,"count"])
neg_dom4 <- as.numeric(longevity_consequence_long[8,"count"])
pos_dom5 <- as.numeric(longevity_consequence_long[9,"count"])
neg_dom5 <- as.numeric(longevity_consequence_long[10,"count"])
pos_dom6 <- as.numeric(longevity_consequence_long[11,"count"])
neg_dom6 <- as.numeric(longevity_consequence_long[12,"count"])
pos_dom7 <- as.numeric(longevity_consequence_long[13,"count"])
neg_dom7 <- as.numeric(longevity_consequence_long[14,"count"])
pos_dom8 <- as.numeric(longevity_consequence_long[15,"count"])
neg_dom8 <- as.numeric(longevity_consequence_long[16,"count"])
pos_dom9 <- as.numeric(longevity_consequence_long[17,"count"])
neg_dom9 <- as.numeric(longevity_consequence_long[18,"count"])
pos_dom10 <- as.numeric(longevity_consequence_long[19,"count"])
neg_dom10 <- as.numeric(longevity_consequence_long[20,"count"])
pos_dom11 <- as.numeric(longevity_consequence_long[21,"count"])
neg_dom11 <- as.numeric(longevity_consequence_long[22,"count"])
pos_dom12 <- as.numeric(longevity_consequence_long[23,"count"])
neg_dom12 <- as.numeric(longevity_consequence_long[24,"count"])

for (i in 1:12) {
  pos_name <- paste("pos_dom", i, sep = "")
  neg_name <- paste("neg_dom", i, sep = "")
  pos <- get(pos_name)
  neg <- get(neg_name)
  if (!(pos == 0)) {
    L3 <- L3 + annotate("text", x = i, y = 15, label = as.character(pos), size = 5/.pt, color = "black")
  }
  if (!(neg == 0)) {
    L3 <- L3 + annotate("text", x = i, y = pos+neg+25, label = as.character(neg), size = 5/.pt, color = "black")
  }
}
```

### Longevity by biotype

Create new data-frame centered in the biotype:

```{r}
longevity_biotype <- longevity %>%
  group_by(BIOTYPE) %>%
  summarize(total_positive = sum(pleiotropy_sign == "positive"),
            total_negative = sum(pleiotropy_sign == "negative")) %>%
  mutate(total = total_positive + total_negative,
         proportion_positive = total_positive / total,
         proportion_negative = total_negative / total)
```

Change the names for a shorter version:

```{r}
biotype_dict <- c("processed_pseudogene" = "processed\npseudogene",
                  "processed_transcript" = "processed\ntranscript",
                  "transcribed_unprocessed_pseudogene" = "transcribed\nunprocessed_pseudogene",
                  "unitary_pseudogene" = "unitary\npseudogene",
                  "unprocessed_pseudogene" = "unprocessed\npseudogene")

longevity_biotype <- longevity_biotype %>%
  mutate(
    BIOTYPE = if_else(
      BIOTYPE %in% names(biotype_dict),
      biotype_dict[BIOTYPE],
      BIOTYPE
    )
  )
```

Format the data for plotting:

```{r}
longevity_biotype <- longevity_biotype %>%
  arrange(desc(total))
longevity_biotype$order <- c(1:15)

longevity_biotype_long <- longevity_biotype %>%
  pivot_longer(cols = starts_with("total"), names_to = "type", values_to = "count") %>%
  mutate(type = recode(type, total_positive = "Positive", total_negative = "Negative"))

longevity_biotype_long <- longevity_biotype_long[longevity_biotype_long$type != "total", ]
```

Plot **number of pleiotropies by consequence**:

```{r}
L5 <- ggplot(longevity_biotype_long, aes(x = reorder(BIOTYPE, order),
                                            y = count,
                                            fill = type)) +
  geom_bar(stat = "identity") +
  labs(x = "Biotype", y = "Number of pleiotropies") +
  scale_fill_manual(values = c("#990000CC", "#023E8ACC")) +
  theme_bw()+
  theme(text = element_text(size = 7, family = "Helvetica", color = "black"),
        axis.text.x = element_text(angle = 60, hjust = 1, size = 5),
        axis.text.y = element_text(size = 7),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7, margin = margin(0,10,0,0)),
        plot.margin = margin(10,10,2,10),
        panel.grid = element_blank())

pos_dom1 <- as.numeric(longevity_biotype_long[1,"count"])
neg_dom1 <- as.numeric(longevity_biotype_long[2,"count"])
pos_dom2 <- as.numeric(longevity_biotype_long[3,"count"])
neg_dom2 <- as.numeric(longevity_biotype_long[4,"count"])
pos_dom3 <- as.numeric(longevity_biotype_long[5,"count"])
neg_dom3 <- as.numeric(longevity_biotype_long[6,"count"])
pos_dom4 <- as.numeric(longevity_biotype_long[7,"count"])
neg_dom4 <- as.numeric(longevity_biotype_long[8,"count"])
pos_dom5 <- as.numeric(longevity_biotype_long[9,"count"])
neg_dom5 <- as.numeric(longevity_biotype_long[10,"count"])
pos_dom6 <- as.numeric(longevity_biotype_long[11,"count"])
neg_dom6 <- as.numeric(longevity_biotype_long[12,"count"])
pos_dom7 <- as.numeric(longevity_biotype_long[13,"count"])
neg_dom7 <- as.numeric(longevity_biotype_long[14,"count"])
pos_dom8 <- as.numeric(longevity_biotype_long[15,"count"])
neg_dom8 <- as.numeric(longevity_biotype_long[16,"count"])
pos_dom9 <- as.numeric(longevity_biotype_long[17,"count"])
neg_dom9 <- as.numeric(longevity_biotype_long[18,"count"])
pos_dom10 <- as.numeric(longevity_biotype_long[19,"count"])
neg_dom10 <- as.numeric(longevity_biotype_long[20,"count"])
pos_dom11 <- as.numeric(longevity_biotype_long[21,"count"])
neg_dom11 <- as.numeric(longevity_biotype_long[22,"count"])
pos_dom12 <- as.numeric(longevity_biotype_long[23,"count"])
neg_dom12 <- as.numeric(longevity_biotype_long[24,"count"])
pos_dom13 <- as.numeric(longevity_biotype_long[25,"count"])
neg_dom13 <- as.numeric(longevity_biotype_long[26,"count"])
pos_dom14 <- as.numeric(longevity_biotype_long[27,"count"])
neg_dom14 <- as.numeric(longevity_biotype_long[28,"count"])
pos_dom15 <- as.numeric(longevity_biotype_long[29,"count"])
neg_dom15 <- as.numeric(longevity_biotype_long[30,"count"])

for (i in 1:15) {
  pos_name <- paste("pos_dom", i, sep = "")
  neg_name <- paste("neg_dom", i, sep = "")
  pos <- get(pos_name)
  neg <- get(neg_name)
  if (!(pos == 0)) {
    L5 <- L5 + annotate("text", x = i, y = 25, label = as.character(pos), size = 5/.pt, color = "black")
  }
  if (!(neg == 0)) {
    L5 <- L5 + annotate("text", x = i, y = pos+neg+60, label = as.character(neg), size = 5/.pt, color = "black")
  }
}
```

Aggregate plots and save:

```{r}
grid_longevity <- plot_grid(L1, L3, L5,
                      ncol = 1,
                      labels = c("a", "b", "c"),
                      label_fontface = "bold",
                      label_fontfamily = "Helvetica",
                      label_size = 7,
                      align = "hv",
                      rel_widths = c(1, 1))

ggsave("/extended_data_fig_4.png", grid_longevity, height = 180, width = 180, units = "mm", dpi = 400)
```

# Legend

```{r}
legend_plot <- ggplot(
  data.frame(
    group = c("Positive Pleiotropy", "Negative Pleiotropy"),
    x = 1, y = 1
  ),
  aes(x = x, y = y, color = group)
) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c(
      "Positive Pleiotropy" = "#023E8A",
      "Negative Pleiotropy"   = "#990000"
    )
  ) +
  guides(color = guide_legend(
    title = NULL,
    override.aes = list(shape = 15, size = 2.5)
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6, family = "Helvetica", color = "black"),
    legend.key.height = unit(0.60, "cm"),
    legend.key.width  = unit(0.20, "cm"),
    legend.background = element_rect(fill = "white", color = "white", linewidth = 0.3),
    legend.box.background = element_rect(fill = "white", color = "white", linewidth = 0.3),
    legend.margin = margin(2, 2, 2, 2)
  )

legend_grob <- get_legend(legend_plot)

grid_longevity_legend <- ggdraw(grid_longevity) +
  draw_grob(
    legend_grob,
    x = 0.8, y = 0.80,
    width = 0.20, height = 0.25
  )

grid_longevity_legend
```

Vector file:

```{r}
ggsave(
  filename = "extended_data_fig_4.pdf",
  plot = grid_longevity_legend,
  device = "pdf",
  width = 180,
  height = 180,
  units = "mm",
  useDingbats = FALSE
)
```

## Fertility

Merge the pleiotropy results with the background set. If an RSID has two or more
consequences (due to different gene isoforms for example), then all of them will
appear in the data-frame.

```{r}
fertility <- merge(fertility_pleios, info_background, by = "rsid")
```

Merge with the domains:

```{r}
fertility <- merge(fertility, domains, by = "disease_code")
```

Create a dictionary to map the domain names with new shorter version:

```{r}
fertility$DomainShort <- domain_dict[fertility$Domain]
```

Create new data-frame centered in the disease domains:

```{r}
fertility_domain <- fertility %>%
  group_by(DomainShort) %>%
  summarize(total_positive = sum(pleiotropy_sign == "positive"),
            total_negative = sum(pleiotropy_sign == "negative")) %>%
  mutate(total = total_positive + total_negative,
         proportion_positive = total_positive / total,
         proportion_negative = total_negative / total)
```

Format the data for plotting:

```{r}
fertility_domain <- fertility_domain %>%
  arrange(desc(total))
fertility_domain$order <- c(1:11)

fertility_domain_long <- fertility_domain %>%
  pivot_longer(cols = starts_with("total"), names_to = "type", values_to = "count") %>%
  mutate(type = recode(type, total_positive = "Positive", total_negative = "Negative"))

fertility_domain_long <- fertility_domain_long[fertility_domain_long$type != "total", ]
```

Plot **number of pleiotropies by domain**:

```{r}
F1 <- ggplot(fertility_domain_long, aes(x = reorder(DomainShort, order),
                                            y = count,
                                            fill = type)) +
  geom_bar(stat = "identity") +
  labs(x = "Disease domain", y = "Number of pleiotropies") +
  scale_fill_manual(values = c("#990000CC", "#023E8ACC")) +
  theme_bw()+
  theme(text = element_text(size = 7, family = "Helvetica", color = "black"),
        axis.text.x = element_text(angle = 60, hjust = 1, size = 5),
        axis.text.y = element_text(size = 7),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7, margin = margin(0,20,0,0)),
        plot.margin = margin(10,10,2,10),
        panel.grid = element_blank())

pos_dom1 <- as.numeric(fertility_domain_long[1,"count"])
neg_dom1 <- as.numeric(fertility_domain_long[2,"count"])
pos_dom2 <- as.numeric(fertility_domain_long[3,"count"])
neg_dom2 <- as.numeric(fertility_domain_long[4,"count"])
pos_dom3 <- as.numeric(fertility_domain_long[5,"count"])
neg_dom3 <- as.numeric(fertility_domain_long[6,"count"])
pos_dom4 <- as.numeric(fertility_domain_long[7,"count"])
neg_dom4 <- as.numeric(fertility_domain_long[8,"count"])
pos_dom5 <- as.numeric(fertility_domain_long[9,"count"])
neg_dom5 <- as.numeric(fertility_domain_long[10,"count"])
pos_dom6 <- as.numeric(fertility_domain_long[11,"count"])
neg_dom6 <- as.numeric(fertility_domain_long[12,"count"])
pos_dom7 <- as.numeric(fertility_domain_long[13,"count"])
neg_dom7 <- as.numeric(fertility_domain_long[14,"count"])
pos_dom8 <- as.numeric(fertility_domain_long[15,"count"])
neg_dom8 <- as.numeric(fertility_domain_long[16,"count"])
pos_dom9 <- as.numeric(fertility_domain_long[17,"count"])
neg_dom9 <- as.numeric(fertility_domain_long[18,"count"])
pos_dom10 <- as.numeric(fertility_domain_long[19,"count"])
neg_dom10 <- as.numeric(fertility_domain_long[20,"count"])
pos_dom11 <- as.numeric(fertility_domain_long[21,"count"])
neg_dom11 <- as.numeric(fertility_domain_long[22,"count"])

for (i in 1:11) {
  pos_name <- paste("pos_dom", i, sep = "")
  neg_name <- paste("neg_dom", i, sep = "")
  pos <- get(pos_name)
  neg <- get(neg_name)
  if (!(pos == 0)) {
    F1 <- F1 + annotate("text", x = i, y = 5, label = as.character(pos), size = 5/.pt, color = "black")
  }
  if (!(neg == 0)) {
    F1 <- F1 + annotate("text", x = i, y = pos+neg+5, label = as.character(neg), size = 5/.pt, color = "black")
  }
}
```

### Fertility by functional consequence

Create new data-frame centered in the functional consequence:

```{r}
fertility_consequence <- fertility %>%
  group_by(Consequence) %>%
  summarize(total_positive = sum(pleiotropy_sign == "positive"),
            total_negative = sum(pleiotropy_sign == "negative")) %>%
  mutate(total = total_positive + total_negative,
         proportion_positive = total_positive / total,
         proportion_negative = total_negative / total)
```

Change the names for a shorter version:

```{r}
conseq_dict <- c("intron_variant,non_coding_transcript_variant" = "intron_variant\nnon_coding_transcript",
                  "non_coding_transcript_exon_variant" = "non_coding_transcript\nexon_variant")

fertility_consequence <- fertility_consequence %>%
  mutate(
    Consequence = if_else(
      Consequence %in% names(conseq_dict),
      conseq_dict[Consequence],
      Consequence
    )
  )
```

Format the data for plotting:

```{r}
fertility_consequence <- fertility_consequence %>%
  arrange(desc(total))
fertility_consequence$order <- c(1:9)

fertility_consequence_long <- fertility_consequence %>%
  pivot_longer(cols = starts_with("total"), names_to = "type", values_to = "count") %>%
  mutate(type = recode(type, total_positive = "Positive", total_negative = "Negative"))

fertility_consequence_long <- fertility_consequence_long[fertility_consequence_long$type != "total", ]
```

Plot **number of pleiotropies by consequence**:

```{r}
F3 <- ggplot(fertility_consequence_long, aes(x = reorder(Consequence, order),
                                            y = count,
                                            fill = type)) +
  geom_bar(stat = "identity") +
  labs(x = "Functional consequence", y = "Number of pleiotropies") +
  scale_fill_manual(values = c("#990000CC", "#023E8ACC")) +
  theme_bw()+
  theme(text = element_text(size = 5, family = "Helvetica", color = "black"),
        axis.text.x = element_text(angle = 60, hjust = 1, size = 5),
        axis.text.y = element_text(size = 7),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7, margin = margin(0,20,0,0)),
        plot.margin = margin(10,10,2,10),
        panel.grid = element_blank())

pos_dom1 <- as.numeric(fertility_consequence_long[1,"count"])
neg_dom1 <- as.numeric(fertility_consequence_long[2,"count"])
pos_dom2 <- as.numeric(fertility_consequence_long[3,"count"])
neg_dom2 <- as.numeric(fertility_consequence_long[4,"count"])
pos_dom3 <- as.numeric(fertility_consequence_long[5,"count"])
neg_dom3 <- as.numeric(fertility_consequence_long[6,"count"])
pos_dom4 <- as.numeric(fertility_consequence_long[7,"count"])
neg_dom4 <- as.numeric(fertility_consequence_long[8,"count"])
pos_dom5 <- as.numeric(fertility_consequence_long[9,"count"])
neg_dom5 <- as.numeric(fertility_consequence_long[10,"count"])
pos_dom6 <- as.numeric(fertility_consequence_long[11,"count"])
neg_dom6 <- as.numeric(fertility_consequence_long[12,"count"])
pos_dom7 <- as.numeric(fertility_consequence_long[13,"count"])
neg_dom7 <- as.numeric(fertility_consequence_long[14,"count"])
pos_dom8 <- as.numeric(fertility_consequence_long[15,"count"])
neg_dom8 <- as.numeric(fertility_consequence_long[16,"count"])
pos_dom9 <- as.numeric(fertility_consequence_long[17,"count"])
neg_dom9 <- as.numeric(fertility_consequence_long[18,"count"])

for (i in 1:9) {
  pos_name <- paste("pos_dom", i, sep = "")
  neg_name <- paste("neg_dom", i, sep = "")
  pos <- get(pos_name)
  neg <- get(neg_name)
  if (!(pos == 0)) {
    F3 <- F3 + annotate("text", x = i, y = 5, label = as.character(pos), size = 5/.pt, color = "black")
  }
  if (!(neg == 0)) {
    F3 <- F3 + annotate("text", x = i, y = pos+neg+12, label = as.character(neg), size = 5/.pt, color = "black")
  }
}
```

### Fertility by biotype

Create new data-frame centered in the biotype:

```{r}
fertility_biotype <- fertility %>%
  group_by(BIOTYPE) %>%
  summarize(total_positive = sum(pleiotropy_sign == "positive"),
            total_negative = sum(pleiotropy_sign == "negative")) %>%
  mutate(total = total_positive + total_negative,
         proportion_positive = total_positive / total,
         proportion_negative = total_negative / total)
```

Change the names for a shorter version:

```{r}
biotype_dict <- c("processed_pseudogene" = "processed\npseudogene",
                  "processed_transcript" = "processed\ntranscript",
                  "transcribed_processed_pseudogene" = "transcribed\nprocessed_pseudogene")

fertility_biotype <- fertility_biotype %>%
  mutate(
    BIOTYPE = if_else(
      BIOTYPE %in% names(biotype_dict),
      biotype_dict[BIOTYPE],
      BIOTYPE
    )
  )
```

Format the data for plotting:

```{r}
fertility_biotype <- fertility_biotype %>%
  arrange(desc(total))
fertility_biotype$order <- c(1:11)

fertility_biotype_long <- fertility_biotype %>%
  pivot_longer(cols = starts_with("total"), names_to = "type", values_to = "count") %>%
  mutate(type = recode(type, total_positive = "Positive", total_negative = "Negative"))

fertility_biotype_long <- fertility_biotype_long[fertility_biotype_long$type != "total", ]
```

Plot **number of pleiotropies by consequence**:

```{r}
F5 <- ggplot(fertility_biotype_long, aes(x = reorder(BIOTYPE, order),
                                            y = count,
                                            fill = type)) +
  geom_bar(stat = "identity") +
  labs(x = "Biotype", y = "Number of pleiotropies") +
  scale_fill_manual(values = c("#990000CC", "#023E8ACC")) +
  theme_bw()+
  theme(text = element_text(size = 5, family = "Helvetica", color = "black"),
        axis.text.x = element_text(angle = 60, hjust = 1, size = 5),
        axis.text.y = element_text(size = 7),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7, margin = margin(0,20,0,0)),
        plot.margin = margin(10,10,2,10),
        panel.grid = element_blank())

pos_dom1 <- as.numeric(fertility_biotype_long[1,"count"])
neg_dom1 <- as.numeric(fertility_biotype_long[2,"count"])
pos_dom2 <- as.numeric(fertility_biotype_long[3,"count"])
neg_dom2 <- as.numeric(fertility_biotype_long[4,"count"])
pos_dom3 <- as.numeric(fertility_biotype_long[5,"count"])
neg_dom3 <- as.numeric(fertility_biotype_long[6,"count"])
pos_dom4 <- as.numeric(fertility_biotype_long[7,"count"])
neg_dom4 <- as.numeric(fertility_biotype_long[8,"count"])
pos_dom5 <- as.numeric(fertility_biotype_long[9,"count"])
neg_dom5 <- as.numeric(fertility_biotype_long[10,"count"])
pos_dom6 <- as.numeric(fertility_biotype_long[11,"count"])
neg_dom6 <- as.numeric(fertility_biotype_long[12,"count"])
pos_dom7 <- as.numeric(fertility_biotype_long[13,"count"])
neg_dom7 <- as.numeric(fertility_biotype_long[14,"count"])
pos_dom8 <- as.numeric(fertility_biotype_long[15,"count"])
neg_dom8 <- as.numeric(fertility_biotype_long[16,"count"])
pos_dom9 <- as.numeric(fertility_biotype_long[17,"count"])
neg_dom9 <- as.numeric(fertility_biotype_long[18,"count"])
pos_dom10 <- as.numeric(fertility_biotype_long[19,"count"])
neg_dom10 <- as.numeric(fertility_biotype_long[20,"count"])
pos_dom11 <- as.numeric(fertility_biotype_long[21,"count"])
neg_dom11 <- as.numeric(fertility_biotype_long[22,"count"])

for (i in 1:11) {
  pos_name <- paste("pos_dom", i, sep = "")
  neg_name <- paste("neg_dom", i, sep = "")
  pos <- get(pos_name)
  neg <- get(neg_name)
  if (!(pos == 0)) {
    F5 <- F5 + annotate("text", x = i, y = 10, label = as.character(pos), size = 5/.pt, color = "black")
  }
  if (!(neg == 0)) {
    F5 <- F5 + annotate("text", x = i, y = pos+neg+25, label = as.character(neg), size = 5/.pt, color = "black")
  }
}
```

Aggregate plots and save:

```{r}
grid_fertility <- plot_grid(F1, F3, F5,
                      ncol = 1,
                      labels = c("a", "b", "c"),
                      label_fontface = "bold",
                      label_fontfamily = "Helvetica",
                      label_size = 7,
                      align = "hv",
                      rel_widths = c(1, 1))

ggsave("extended_data_fig_5.png", grid_fertility, height = 170, width = 170, units = "mm", dpi = 400)
```

# Legend

```{r}
legend_plot <- ggplot(
  data.frame(
    group = c("Positive Pleiotropy", "Negative Pleiotropy"),
    x = 1, y = 1
  ),
  aes(x = x, y = y, color = group)
) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c(
      "Positive Pleiotropy" = "#023E8A",
      "Negative Pleiotropy"   = "#990000"
    )
  ) +
  guides(color = guide_legend(
    title = NULL,
    override.aes = list(shape = 15, size = 2.5)
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6, family = "Helvetica", color = "black"),
    legend.key.height = unit(0.60, "cm"),
    legend.key.width  = unit(0.20, "cm"),
    legend.background = element_rect(fill = "white", color = "white", linewidth = 0.3),
    legend.box.background = element_rect(fill = "white", color = "white", linewidth = 0.3),
    legend.margin = margin(2, 2, 2, 2)
  )

legend_grob <- get_legend(legend_plot)

grid_fertility_legend <- ggdraw(grid_fertility) +
  draw_grob(
    legend_grob,
    x = 0.8, y = 0.80,
    width = 0.20, height = 0.25
  )

grid_fertility_legend
```

Vector file:

```{r}
ggsave(
  filename = "extended_data_fig_5.pdf",
  plot = grid_fertility_legend,
  device = "pdf",
  width = 180,
  height = 180,
  units = "mm",
  useDingbats = FALSE
)
```
